---
title: "Exploratory Data Analysis"
author: "Isaac Baca"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This project demonstrates a variety of data exploration techniques using three different data sets:

The *comics* data set which contains information on all comic characters that have been introduced by DC and Marvel is used to explore categorical data.

The *cars* data set which contains information on all new models of cars for sale in the US in a certain year is used to explore numerical data.

The *gapminder* data set which contains information on demographic data in various countries throughout the world is used to explore numerical summaries.

---

# Table of contents
1. [Exploring Categorical Data](#one)
2. [Exploring Numerical Data](#two)
3. [Numerical Summaries](#three)

------------------------------------------------------------------------

# Load all necessary packages.

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```

------------------------------------------------------------------------

<a name="one"><a/>

# Exploring Categorical Data.

#### Import the comics data set and view the structure.

-   The comics data set has information on all comic characters that have been introduced by DC and Marvel.

```{r echo=FALSE}
comics = read.csv("/Volumes/Vandy Main/Data Camp/R/Exploratory Data Analysis/comics.csv")
```

```{r}
str(comics)
```

------------------------------------------------------------------------

#### Some of the variables have a `chr` variable type. We need to change the type to `Factor`.

```{r}
comics[sapply(comics, is.character)] <- lapply(comics[sapply(comics, is.character)], 
                                       as.factor)
str(comics)
```
`str(comics)` shows us that all `chr` variable types have now been changed to `Factor`.

------------------------------------------------------------------------

#### Check levels of align.

We can see that there are four possible levels that the align variable can take.

```{r}
levels(comics$align)
```

------------------------------------------------------------------------

#### Check the levels of gender.

There are three possible levels that the gender variable can take.

```{r}
levels(comics$gender)
```

------------------------------------------------------------------------

#### Create a 2-way contingency table of the two variables.

I will name this table `tab` for future use.

```{r}
tab = table(comics$align, comics$gender)
print(tab)
```

------------------------------------------------------------------------

#### The contingency table from the last exercise revealed that there are some levels that have very low counts. To simplify the analysis, we will drop these levels.

-   `filter(align != "Reformed Criminals")` will filter the data set for everything that is not "Reformed Criminals".

-   `droplevels()` will drop unused levels from a factor.

```{r}
comics <- comics %>%
  filter(align != "Reformed Criminals") %>% 
  droplevels()
```

------------------------------------------------------------------------

#### The `align` variable now has three levels.

```{r}
str(comics)
```

------------------------------------------------------------------------

#### We can represent a contingency table graphically for ease of interpretation using a side-by-side bar chart.

-   First, we create a side-by-side bar chart with `align` on the x-axis and `gender` as the fill aesthetic.

-   Passing the argument `position = "dodge"` to `geom_bar()` says that you want a side-by-side (i.e. not stacked) bar chart.

```{r}
ggplot(comics, aes(x = align, fill = gender)) + 
  geom_bar(position = "dodge")
```

------------------------------------------------------------------------

#### We can switch the x-axis and y-axis variables

-   Create a side-by-side bar chart with `gender` on the x-axis and `align` as the fill aesthetic.
-   `theme(axis.text.x = element_text(angle = 90))` rotates the axis lables 90 degrees to enhance readability.

```{r}
ggplot(comics, aes(x = gender, fill = align)) + 
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 90))
```

------------------------------------------------------------------------

#### Obtain a conditional proportion table.

-   The second argument in `prop.table()` specifies whether to condition on the rows or the columns.
-   `1` conditions on the rows, `2` conditions on the columns.

```{r}
prop.table(tab, 1)
prop.table(tab, 2)
```

---

#### Counts vs Proportions represented graphically with bar charts.

-   First, we create a stacked bar chart of `gender` *counts* with `align` on the x-axis.

```{r}
ggplot(comics, aes(x = align, fill = gender)) +
  geom_bar()
```

---

#### The *proportions* tell a differenct story than the counts.

-   Next, we create a stacked bar chart of `gender` *proportions* with `align` on the x-axis.

```{r}
ggplot(comics, aes(x = align, fill = gender)) + 
  geom_bar(position = "fill") +
  ylab("proportion")
```

---

#### View the distribution of alignment of all superheroes.

-   First, we change the order of the levels in `align` to improve the interpretability of the plot by passing a vector to the second argument of `factor()`.
-   Then, we construct a bar chart that represents the distribution of alignment of all superheroes.

```{r}
comics$align <- factor(comics$align, levels = c("Bad", "Neutral", "Good"))

ggplot(comics, aes(x = align)) + 
  geom_bar()
```

---

#### Conditional distributions.

-   We can further break down the distribution of alignment based on gender.
-   We can use `facet_wrap(~ gender)` to facet by gender.

```{r}
ggplot(comics, aes(x = align)) + 
  geom_bar() +
  facet_wrap(~ gender)
```

---

<a name="two"><a/>

# Exploring Numerical Data.

#### Import the comics data set and view the structure.

-   The `cars` data set records characteristics on all of the new models of cars for sale in the US in a certain year.

```{r}
cars = read.csv("/Volumes/Vandy Main/Data Camp/R/Exploratory Data Analysis/cars04.csv")
str(cars)
```

---

#### Plot a histogram of `city_mpg` faceted by `suv`, a logical variable indicating whether the car is an SUV or not.

```{r}
ggplot(cars, aes(x = city_mpg)) +
  geom_histogram() +
  facet_wrap(~ suv)
```

---

#### Explore the relationship between the milage of a car and the size of the engine.

-   First, we filter the data set for cars with 4, 6, or 8 cylinders and save the result as `common_cyl`.

```{r}
common_cyl <- filter(cars, ncyl == 4 | ncyl == 6 | ncyl == 8)
```

---

#### Create side-by-side box plots of `city_mpg` separated out by `ncyl`.

```{r}
ggplot(common_cyl, aes(x = as.factor(ncyl), y = city_mpg)) +
  geom_boxplot()
```

---

#### Create overlaid density plots of `city_mpg` colored by `ncyl`.

```{r}
ggplot(common_cyl, aes(x= city_mpg, fill = as.factor(ncyl))) +
  geom_density(alpha = .3)
```

---

#### Create a histogram of the distribution of `horsepwr` across all cars.

```{r}
cars %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram() +
  ggtitle("Distribution of Horsepower")
```

---

#### Create a second histogram of the distribution of `horsepwr`, but only for those cars that have an `msrp` less than \$25,000.

-   Set the `xlim()` and `ylim()` to match the previous histogram.

```{r}
cars %>% 
  filter(msrp < 25000) %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram() +
  ylim(c(0, 50)) +
  xlim(c(90, 550)) +
  ggtitle("Distribution of Horsepower for Affordable Cars")
```

---

#### View how the histogram changes as the binwidth changes.

-   First, we set `binwidth = 3`.
-   Next, we set `binwidth = 30`.
-   Finally, we set `binwidth = 60`.

```{r}
cars %>%
  ggplot(aes(horsepwr)) +
  geom_histogram(binwidth = 3) +
  ggtitle("binwidth = 3")


cars %>%
  ggplot(aes(horsepwr)) +
  geom_histogram(binwidth = 30) +
  ggtitle("binwidth = 30")


cars %>%
  ggplot(aes(horsepwr)) +
  geom_histogram(binwidth = 60) +
  ggtitle("binwidth = 60")
```

---

#### In addition to indicating the center and spread of a distribution, a box plot provides a graphical means to detect outliers.

-   I construct a box plot of `msrp` (manufacturer's suggested retail price) to detect if there are unusually expensive or cheap cars.

```{r}
cars %>%
  ggplot(aes(x = 1, y = msrp)) +
  geom_boxplot()
```

---

#### I exclude the largest 3-5 outliers by filtering the rows to retain cars less than \$100,000.

-   This new data set will be called `cars_no_out`.

```{r}
cars_no_out <- cars %>%
  filter(msrp < 100000)
```

---

#### I construct a similar box plot of `msrp` using this reduced dataset.

```{r}
cars_no_out %>%
  ggplot(aes(x = 1, y = msrp)) +
  geom_boxplot()
```

---

#### Both density plots and box plots display the central tendency and spread of the data, but the box plot is more robust to outliers.

-   `city_mpg` can be represented using either a density plot or a box plot.
-   As these plots demonstrate, `city_mpg` contains many outliers. Therefore, a box plot would be the better choice here.

```{r}
cars %>%
  ggplot(aes(x = 1, y = city_mpg)) +
  geom_boxplot() + 
  coord_flip()


cars %>%
  ggplot(aes(x = city_mpg)) +
  geom_density()
```

---

#### Similarly, `width` can be represented using either a density plot or a box plot.

-   In this case, a density plot would be the better choice.

```{r}
cars %>%
  ggplot(aes(x = 1, y = width)) +
  geom_boxplot() + 
  coord_flip()


cars %>%
  ggplot(aes(x = width)) +
  geom_density()
```

---

#### We can look at several conditional distributions at the same time by using `facet_grid`.

-   `common_cyl` contains only cars with 4, 6, or 8 cylinders.
-   I use `common_cyl` to create a histogram of `hwy_mpg`.
-   I grid-facet the plot rowwise by `ncyl` and columnwise by `suv`.

```{r}
common_cyl %>%
  ggplot(aes(x = hwy_mpg)) +
  geom_histogram() +
  facet_grid(ncyl ~ suv, labeller = label_both) +
  ggtitle("Highway MPG for 4, 6, and 8 cyl SUVs and non-SUVs")
```

---

<a name="three"><a/>

# Numerical Summaries.

#### Import the gapminder data set and view the structure.

-   The `gapminder` data set tracks demographic data in various countries throughout the world over time.

```{r}
gapminder = read.csv("/Volumes/Vandy Main/Data Camp/R/Exploratory Data Analysis/gapminder.csv")
str(gapminder)
```

---

#### Some of the variables have a `chr` variable type. We need to change the type to `Factor`.

-   I check the structure of the data set again to make sure all `chr` variables were changed to `Factor`.

```{r}
gapminder[sapply(gapminder, is.character)] <- lapply(gapminder[sapply(gapminder, is.character)], 
                                               as.factor)
str(gapminder)
```

---

#### I start by creating a dataset called `gap2007` that contains only data from the year 2007.

```{r}
gap2007 <- filter(gapminder, year == 2007)
```

---

#### Using `gap2007`, I calculate the mean and median life expectancy for each continent.

```{r}
gap2007 %>%
  group_by(continent) %>%
  summarize(mean(lifeExp), median(lifeExp))
```

---

#### I confirm the trends that I see in the medians by generating side-by-side box plots of life expectancy for each continent.

-   This also enables us to view the spread of life expectancy for each continent, as well as any potential outliers.

```{r}
gap2007 %>%
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot()
```

---

#### If its unclear whether the distribution of the data is symmetric or skewed, it's a good idea to consider a robust measure like **IQR** in addition to the usual measures of **variance** or **standard deviation**.

-   For each continent in `gap2007`, I summarize the life expectancies using `sd()`, the `IQR()`, and the count of countries, `n()`.

```{r}
gap2007 %>%
  group_by(continent) %>%
  summarize(sd(lifeExp), IQR(lifeExp), n())
```

---

#### I compare the spread of these distributions by constructing overlaid density plots of life expectancy broken down by continent.

```{r}
gap2007 %>%
  ggplot(aes(x = lifeExp, fill = continent)) +
  geom_density(alpha = 0.3)
```

---

#### I want to determine the most appropriate measures of center and spread for the distribution of life expectancy in the Americas.

-   To view the spread, I create a density plot for life expectancy in the Americas.

```{r}
gap2007 %>%
  filter(continent == "Americas") %>%
  ggplot(aes(x = lifeExp)) +
  geom_density() 
```

-   We can see that the distribution is farily normal.
-   Therefore, I use `mean()` and `sd()` as measures of center and spread.

```{r}
gap2007 %>%
  filter(continent == "Americas") %>%
  summarize(mean(lifeExp), sd(lifeExp))
```

---

#### I want to determine the appropriate measures of center and spread for the distribution of country *populations* across the entire `gap2007` data set.

-   To view the spread, I create a density plot for country level populations.

```{r}
gap2007 %>%
  ggplot(aes(x = pop)) +
  geom_density() 
```

-   We can see that the density plot has a strong skew.
-   Therefore, I use `median()` and `IQR()` as measures of center and spread.

```{r}
gap2007 %>%
  summarize(median(pop), IQR(pop))
```

---

#### Highly skewed distributions can make it very difficult to learn anything from a visualization. Transformations can be helpful in revealing the more subtle structure.

-   As the last plot demonstrated, the population variable is highly skewed.

```{r}
gap2007 %>%
  ggplot(aes(x = pop)) +
  geom_density()
```

---

#### The population variable can be transformed using the natural logarithm funciton, `log()`.

-   I take the transformed variable and add it to the `gap2007` data frame using `mutate(log_pop = log(pop))`.

```{r}
gap2007 <- gap2007 %>%
  mutate(log_pop = log(pop))
str(gap2007)
```

---

#### I create a density plot of the transformed population variable.

-   Using the transformed population variable makes for a much better visualization.

```{r}
gap2007 %>%
  ggplot(aes(x = log_pop)) +
  geom_density()
```

---

#### The box plot below visualizes life expenctancies of countries in Asia.

-   There is one clear outlier: a country with a notably low life expectancy.

```{r}
gap2007 %>%
  filter(continent == "Asia") %>%
  ggplot(aes(x = continent, y = lifeExp)) + 
  geom_boxplot()
```

---

#### I use `min(gap2007_Asia$lifeExp)` to find the outlier.

```{r}
gap2007_Asia = gap2007 %>%
  filter(continent == "Asia")

  min(gap2007_Asia$lifeExp)

```

---

#### I find the observation where life expectancy is 43.828.

```{r}
gap2007_Asia %>%
    filter(lifeExp == 43.828)
```

---

#### I create a new variable called `is_outlier` that is `TRUE` for countries with life expectancy less than 50.

-   This variable is only true for Afghanistan.

```{r}
gap_asia <- gap2007 %>%
 filter(continent == "Asia") %>%
 mutate(is_outlier = lifeExp < 50)
head(gap_asia)
```

---

#### I filter `gap_asia` to remove all outliers, then create another box plot of the remaining life expectancies.

```{r}
gap_asia %>%
 filter(!is_outlier) %>%
 ggplot(aes(x = 1, y = lifeExp)) +
 geom_boxplot()
```
